---
title: "Reproducible Research Assignment 1"
author: "James Jackson"
date: "October 30, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown Doc

This is my R Markdown document that answers the questopns asked.

### Loading and processing data
I unzip the file once then ignore in future runs

1.& 2.
```{r echo=TRUE}
setwd("C:/Users/Sophy/Desktop/JJ Coursera/Reproducible Research/RepData_PeerAssessment1")

#unzip("activity.zip")

activityData <- read.csv("activity.csv",header=T)
```


### What is mean total number of steps taken per day?

1.
```{r echo=TRUE}
totalStepsperdayData <- aggregate(activityData$steps~activityData$date,activityData,FUN = sum, 
              na.action = na.omit)
```

2.
```{r echo=TRUE}
hist(totalStepsperdayData[,2], xlab = "Number of Steps per day", main = "Histogram of Steps", breaks = 10)
```

3.
```{r echo=TRUE}
print(paste("The mean total number of steps per day is:", round(mean(totalStepsperdayData[,2]),digits = 2)))
print(paste("The median total number of steps per day is:", round(median(totalStepsperdayData[,2]),digits=2)))
```


### What is the average daily activity pattern?

1.
```{r echo=TRUE}
averageStepsperintData <- aggregate(activityData$steps~activityData$interval,activityData,FUN = mean,na.action = na.omit)
plot(averageStepsperintData[,1],averageStepsperintData[,2],xlab="Interval (start times)",ylab="Average Steps", main="Average Steps per interval (excl NA data)",type="l")
```

2.
```{r echo=TRUE}
rowIntMaxStep <- averageStepsperintData[which.max(averageStepsperintData[,2]),]
print(paste("Starting time of max average steps (5min interval) is:", rowIntMaxStep[1]))
```


### Imputing missing values

1.
```{r echo=TRUE}
sum(is.na(activityData$steps))
```

2.
This method is about the most basic I can think of.  It replaces every NA value with the average number of steps per interval, across all days (and intervals)
```{r echo=TRUE}
averageSteps <- mean(averageStepsperintData[,2])
activityData[is.na(activityData)] <- averageSteps
```

3.
```{r echo=TRUE}
noNAtotalStepsperdayData <- aggregate(activityData$steps~activityData$date,activityData,FUN = sum, na.action = na.omit)
```

4.
```{r echo=TRUE}
hist(noNAtotalStepsperdayData[,2], xlab = "Number of Steps per day", main = "Histogram of Steps (replaced NAs)", breaks = 10)

print(paste("The mean total number of steps per day is", round(mean(noNAtotalStepsperdayData[,2]),digits = 2)))
print(paste("The median total number of steps per day is", round(median(noNAtotalStepsperdayData[,2]),digits = 2)))
```
Comments: My method of replacing the NA values with the average accross all intervals has had the predicted effect. The mean and median are now the same as I've added more values at the middle. This is shown in the hist plots as all frequencies remain the same, except for the middle bar where the frequency has increased due to the replacement of the NA values being assigned the middle value.

### Are there differences in activity patterns between weekdays and weekends?

1.
```{r echo=TRUE}
#leave NA values out, as in original
activityData$day <- weekdays(as.Date(activityData$date,'%Y-%m-%d'))
weekdaysref <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
activityData$wday <- factor((activityData$day %in% weekdaysref), levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))

weekdayData <- activityData[activityData$wday == "weekday",]
weekendData <- activityData[activityData$wday == "weekend",]

averageStepsperintWeekdayData <- aggregate(weekdayData$steps~weekdayData$interval,weekdayData,FUN = mean,na.action = na.omit)
averageStepsperintWeekendData <- aggregate(weekendData$steps~weekendData$interval,weekendData,FUN = mean,na.action = na.omit)
```

2.
```{r echo=TRUE}
par(mfrow = c(2,1))
fig.width=7
fig.height=7
plot(averageStepsperintWeekdayData[,1],averageStepsperintWeekdayData[,2],main="Weekday (excl NA data)",type="l",ylim = c(0,200),xlab = "", ylab="# Steps")
plot(averageStepsperintWeekendData[,1],averageStepsperintWeekendData[,2],xlab="Interval (start times)", main="Weekend (excl NA data)",type="l",ylim = c(0,200),ylab="# Steps")
```
